The Three-HITs Theorem

### 2017-03-29 Paper/Paper.tex
Fix `:`

For example, if $\nat$ is the type of natural numbers, the expression
$(X + \nat) \times X$ is a polynomial expression in~$X$. We write $P[X]$ to indicate a
polynomial in variable $X$. Given a type $A$, we may substutute $A$ for $X$ to obtain a
type $P[A]$, as follows:
%
\begin{align*}
  X[A] &\defeq A, &
  (P_1 \times P_2)[A] &\defeq P_1[A] \times P_2[A], \\
  C[A] &\defeq C, &
  (P_1 + P_2)[A] &\defeq P_1[A] + P_2[A].
\end{align*}
%
In a similar fashion a map $f \> A \to B$ is transformed to a map $P[f] : P[A] \to P[B]$:
%
\begin{align*}
  X[f] &\defeq f, &
  (P_1 \times P_2)[f] \, (x,y) &\defeq (P_1[f]\, x, P_2[f] \, y), \\
  C[f] &\defeq \idmap{C}, &
  (P_1 + P_2)[f](\inl{x}) &\defeq \inl{(P_1[f]\,x)}, \\
  & &
  (P_1 + P_2)[f](\inr{y}) &\defeq \inr{(P_2[f]\,y)}.
\end{align*}
%
Somewhat less obviously, a polynomial acts on a type family $B : A \to \Type$ to
give its lifting $\lift{P}[B] : P[A] \to \Type$,
%
\begin{align*}
  \lift{X}[B] \, u &\defeq B\,u, &
  \lift{P_1 \times P_2}[B] \, u \defeq (\lift{P_1}[B]\,u) \times (\lift{P_2}[B]\,u), \\
  \lift{C}[B] \, u &\defeq C, &
  \lift{P_1 + P_2}[B] \, (\inl{u})\defeq \lift{P_1}[B]\,u, \\
  & &
  \lift{P_1 + P_2}[B] \, (\inr{u})\defeq \lift{P_2}[B]\,u,
\end{align*}
%
while a map $f \> \dprod{x \> A} B_1\,x \to B_2\,x$ lifts to
$\lift{P}[f] \> \dprod{u : P[A]} \lift{P}[B_1]\,u \to \lift{P}[B_2]\,u$:
%
\begin{align*}
  \lift{X}[f] \, u &\defeq f \, u &
  \lift{P_1 \times P_2}[f] \, (u_1, u_2) &\defeq (\lift{P_1}[f]\,u_1, \lift{P_2}[f]\,u_2) \\
  \lift{C}[f] \, u &\defeq u &
  \lift{P_1 + P_2}[f] \, (\inl{u})\defeq \lift{P_1}[f]\,u, \\
  & &
  \lift{P_1 + P_2}[f] \, (\inr{u})\defeq \lift{P_2}[f]\,u.
\end{align*}
%
There is of course a strong analogy between the action of $P[X]$ and a polynomial functor,
but we hesitate to call $P[X]$ a functor as there is no category to speak of, and in any
case the manipulations we have just described are syntactic.

Next we define the notion of a constructor term.
\begin{definition}
\label{def:constructor-term}
Given is a function $c : A \> T \rightarrow T$ with $A$ polynomial and $T$ a type.
Then we say $t$ is a \emph{constructor term} over $c$ if we can find polynomials $F$ and $G$ such that $x : F \> T \vdash t : G \> T$ can be derived using the following rules.
\begin{gather*}
        \AxiomC{$t : B$}
        \AxiomC{$T$ does not occur in $B$}
        \BinaryInfC{$x:F \> T \vdash t : B$}
        \bottomAlignProof
        \DisplayProof 
         \quad
        \AxiomC{}
        \UnaryInfC{$x : F \> T \vdash x : F \> T$}
        \bottomAlignProof
        \DisplayProof
        \quad
        \AxiomC{$x : F \> T \vdash r : A \> T$}
        \UnaryInfC{$x : F \> T \vdash c \> r : T$}
        \bottomAlignProof
        \DisplayProof
        \\[7pt]
        \AxiomC{$j \in \{1, 2\}$}
        \AxiomC{$x : F \> T \vdash r : G_1 \> T \times G_2 \> T$}
        \BinaryInfC{$x : F \> T \vdash\pi_j \> r : G_j \> T$}
        \bottomAlignProof
        \DisplayProof
        \quad
        \AxiomC{$j \in \{1, 2\}$}
        \AxiomC{$x : F \> T \vdash r_j : G_j$}
        \BinaryInfC{$x : F \> T \vdash (r_1, r_2) : G_1 \> T\times G_2 \> T$}
        \bottomAlignProof
        \DisplayProof
        \\[7pt]
        \AxiomC{$j \in \{1, 2\}$}
        \AxiomC{$x : F \> T \vdash r : G_j \> T$}
        \BinaryInfC{$x : F \> T : \inn_j \> r : G_1 \> T + G_2 \> T$}
        \bottomAlignProof
        \DisplayProof
\end{gather*}
\end{definition}

@@

Before we can give the rules for higher inductive types, we need to define the lift of a constructor term.
\begin{definition}
Given is a constructor $c : A \> H \rightarrow H$, a type family $Y : H \rightarrow \Type$, and a term $f : \prod (x: A \> H), \bar{A} \> Y \> x \rightarrow  Y(c \> x)$.
For a constructor term $F \> H \vdash r : G \> H$ we define the \emph{lift $\dak{r}$ of $r$} with type 
$
x : F \> H, h_x : \bar{F} \> Y \> x \vdash \dak{r} : \bar{G} \> Y \> r
$
by induction in $r$ as follows.
        \begin{align*}
                \dak{t} & := t
                & \dak{x} & := h_x &
                \dak{c_i \> r} & := f_i \> r \> \dak{r}\\
                \dak{\pi_j \> r} & := \pi_j \> \dak{r} &
                \dak{(r_1, r_2)} & := (\dak{r_1}, \dak{r_2})
                & \dak{\inn_j \> r} & := \dak{r}
        \end{align*}
\end{definition}

>>>>

For example, if $\nat$ is the type of natural numbers, the expression
$(X + \nat) \times X$ is a polynomial expression in~$X$. We write $P[X]$ to indicate a
polynomial in variable $X$. Given a type $A$, we may substutute $A$ for $X$ to obtain a
type $P[A]$, as follows:
%
\begin{align*}
  X[A] &\defeq A, &
  (P_1 \times P_2)[A] &\defeq P_1[A] \times P_2[A], \\
  C[A] &\defeq C, &
  (P_1 + P_2)[A] &\defeq P_1[A] + P_2[A].
\end{align*}
%
In a similar fashion a map $f : A \to B$ is transformed to a map $P[f] : P[A] \to P[B]$:
%
\begin{align*}
  X[f] &\defeq f, &
  (P_1 \times P_2)[f] \, (x,y) &\defeq (P_1[f]\, x, P_2[f] \, y), \\
  C[f] &\defeq \idmap{C}, &
  (P_1 + P_2)[f](\inl{x}) &\defeq \inl{(P_1[f]\,x)}, \\
  & &
  (P_1 + P_2)[f](\inr{y}) &\defeq \inr{(P_2[f]\,y)}.
\end{align*}
%
Somewhat less obviously, a polynomial acts on a type family $B : A \to \Type$ to
give its lifting $\lift{P}[B] : P[A] \to \Type$,
%
\begin{align*}
  \lift{X}[B] \, u &\defeq B\,u, &
  \lift{P_1 \times P_2}[B] \, u \defeq (\lift{P_1}[B]\,u) \times (\lift{P_2}[B]\,u), \\
  \lift{C}[B] \, u &\defeq C, &
  \lift{P_1 + P_2}[B] \, (\inl{u})\defeq \lift{P_1}[B]\,u, \\
  & &
  \lift{P_1 + P_2}[B] \, (\inr{u})\defeq \lift{P_2}[B]\,u,
\end{align*}
%
while a map $f : \dprod{x \> A} B_1\,x \to B_2\,x$ lifts to
$\lift{P}[f] : \dprod{u : P[A]} \lift{P}[B_1]\,u \to \lift{P}[B_2]\,u$:
%
\begin{align*}
  \lift{X}[f] \, u &\defeq f \, u &
  \lift{P_1 \times P_2}[f] \, (u_1, u_2) &\defeq (\lift{P_1}[f]\,u_1, \lift{P_2}[f]\,u_2) \\
  \lift{C}[f] \, u &\defeq u &
  \lift{P_1 + P_2}[f] \, (\inl{u})\defeq \lift{P_1}[f]\,u, \\
  & &
  \lift{P_1 + P_2}[f] \, (\inr{u})\defeq \lift{P_2}[f]\,u.
\end{align*}
%
There is of course a strong analogy between the action of $P[X]$ and a polynomial functor,
but we hesitate to call $P[X]$ a functor as there is no category to speak of, and in any
case the manipulations we have just described are syntactic.

Next we define the notion of a constructor term.
\begin{definition}
\label{def:constructor-term}
Given a function $c : A \> T \rightarrow T$ with $A$ a polynomial and $T$ a type,
we say $t$ is a \emph{constructor term} over $c$ if we can find polynomials $F$ and $G$ such that $x : F \> T \vdash t : G \> T$ can be derived using the following rules.
\begin{gather*}
        \AxiomC{$t : B$}
        \AxiomC{$T$ does not occur in $B$}
        \BinaryInfC{$x:F \> T \vdash t : B$}
        \bottomAlignProof
        \DisplayProof 
         \quad
        \AxiomC{}
        \UnaryInfC{$x : F \> T \vdash x : F \> T$}
        \bottomAlignProof
        \DisplayProof
        \quad
        \AxiomC{$x : F \> T \vdash r : A \> T$}
        \UnaryInfC{$x : F \> T \vdash c \> r : T$}
        \bottomAlignProof
        \DisplayProof
        \\[7pt]
        \AxiomC{$j \in \{1, 2\}$}
        \AxiomC{$x : F \> T \vdash r : G_1 \> T \times G_2 \> T$}
        \BinaryInfC{$x : F \> T \vdash\pi_j \> r : G_j \> T$}
        \bottomAlignProof
        \DisplayProof
        \quad
        \AxiomC{$j \in \{1, 2\}$}
        \AxiomC{$x : F \> T \vdash r_j : G_j$}
        \BinaryInfC{$x : F \> T \vdash (r_1, r_2) : G_1 \> T\times G_2 \> T$}
        \bottomAlignProof
        \DisplayProof
        \\[7pt]
        \AxiomC{$j \in \{1, 2\}$}
        \AxiomC{$x : F \> T \vdash r : G_j \> T$}
        \BinaryInfC{$x : F \> T \vdash \inn_j \> r : G_1 \> T + G_2 \> T$}
        % shouldn't this be inl/inj instead of in_1/in_2?
        \bottomAlignProof
        \DisplayProof
\end{gather*}
\end{definition}

@@

Before we can give the rules for higher inductive types, we need to define the lift of a constructor term.
\begin{definition}
Given is a constructor $c : A \> H \rightarrow H$, a type family $Y : H \rightarrow \Type$, and a term $f : \prod (x: A \> H), \bar{A} \> Y \> x \rightarrow  Y(c \> x)$.
For a constructor term $x : F \> H \vdash r : G \> H$ we define the \emph{lift $\dak{r}$ of $r$} with type 
$
x : F \> H, h_x : \bar{F} \> Y \> x \vdash \dak{r} : \bar{G} \> Y \> r
$
by induction in $r$ as follows.
        \begin{align*}
                \dak{t} & := t
                & \dak{x} & := h_x &
                \dak{c_i \> r} & := f_i \> r \> \dak{r}\\
                \dak{\pi_j \> r} & := \pi_j \> \dak{r} &
                \dak{(r_1, r_2)} & := (\dak{r_1}, \dak{r_2})
                & \dak{\inn_j \> r} & := \dak{r}
        \end{align*}
\end{definition}
